Approximation as Differentiation

### 2025-05-23 notes/stability.tex
Hopefully illuminating examples of stable and non-stable functions

\begin{definition}
  Let $f : X \to Y$ be a monotone function between posets $X$ and
  $Y$. The function $f$ is \emph{stable} if for all $x \in X$ and
  $y \leq f(x)$, there exists an $x_0 \leq x$ such that $y \leq f(x_0)$ which
  is minimal: for any other $x'_0 \leq x$ such that $y \leq f(x'_0)$ then
  $x_0 \leq x'_0$.
\end{definition}

The intuition behind stability is that for stable functions, we have a
way of determining what minimal approximation of the input is required
for any approximation of the output. The ordering on $X$ and $Y$
describes an ordering of information content between
elements. Monotonicity means that increases in information content on
the input are not lost on the output. Stability means that, if we use
$f$ to derive a result $f(x)$ from $x$, and then we identify some
approximation $y$ of $f(x)$, then there is a unique minimal
approximation of $x$ that gets us as close as possible to $y$.

We can rephrase stability in terms of Galois connections / adjunctions
between principal downsets.

@@

\begin{proof}
  If $f$ is stable, then define a left adjoint
  $g_x : \downset{f(x)} \to \downset{x}$ by setting $g_x(y)$ to be the
  minimal $x_0$ required by stability. (FIXME: this is monotone?). Let
  $x' \leq x$ and $y \leq f(x)$. Then if $g_x(y) \leq x'$, we have
  $y \leq f(g_x(y)) \leq f(x')$ by monotonicity of $f$ and the first
  part of stability. In the other direction, if we have
  $y \leq f(x')$, then by uniqueness we have $g_x(y) \leq x'$.

  TODO: left adjoint implies stability.
\end{proof}

>>>>

Stable functions are an antecedent of Galois slicing that arose in
domain theory as an attempt to capture the idea that any function
describing a computational process must explore its input in a way
that allows identification of minimal parts of the input that are
needed for chosen approximations of the output.

Stability was invented by Berry \cite{berry86} in an attempt to capture
sequentiality, and were also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien} (Chapter 12). Here, we are not
concerned with completeness properties considered in domain theory, so
we start with only partially ordered sets. Stable functions can be
defined as a certain subset of the set of monotone functions between
two partially ordered sets:

\begin{definition}
  Let $f : X \to Y$ be a monotone function between posets $X$ and
  $Y$. The function $f$ is \emph{stable} if for all $x \in X$ and
  $y \leq f(x)$:
  \begin{enumerate}
  \item (\textsc{Existence}) there exists an $x_0 \leq x$ such that $y \leq f(x_0)$, and
  \item (\textsc{Minimality}) for any $x'_0 \leq x$ such that $y \leq f(x'_0)$ then
    $x_0 \leq x'_0$ .
  \end{enumerate}
\end{definition}

\begin{lemma}
  Stable functions are closed under identities and composition.
\end{lemma}

\begin{example}[Stability reveals intensional information]
  To see how stability works, consider the following examples of the
  OR operation on the lifted booleans $\mathbb{B}_\bot$. Two functions
  that are stable are the strict and left-short-circuiting
  ORs\footnote{These clauses are shorthand for the graph of the
    function. They are not to be understood as pattern matching
    clauses in a language like Haskell.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
  \end{displaymath}
  The function $\mathrm{strictOr}$ is stable. For example, for the
  input-output pair $(\mathsf{tt},\mathsf{ff}), \mathsf{tt}$, the
  minimal input that gives this output is exactly
  $(\mathsf{tt}, \mathsf{ff})$. If we take the approximation
  $\bot \leq \mathsf{tt}$ of the output, then the corresponding
  minimal input is $(\bot, \bot)$.

  The function $\mathrm{shortCircuitOR}$ is also stable. For the
  input-output pair $(\mathsf{tt},\mathsf{ff}),\mathsf{tt}$, the
  minimal input that gives this input is $(\mathsf{tt},\bot)$,
  indicating that the presence of $\mathsf{ff}$ in the second argument
  was not necessary to produce this output. As with
  $\mathrm{strictOr}$, the minimal input required to produce the
  output $\bot \leq \mathsf{tt}$ is again $(\bot,\bot)$.

  The fact that these two functions's stability witnesses reveal
  information about how they depend on their input is what we will
  exploit in order to use the idea of stability for slicing.
\end{example}

\begin{example}[Functions that are not stable]
  A function that is not stable is Plotkin's Parallel OR \cite{lcf77},
  which short-circuits in both arguments, returning $\mathsf{tt}$ if
  either argment is $\mathsf{tt}$, even if the other argument is not
  defined:
  \begin{displaymath}
    \begin{array}{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{parallelOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathrm{parallelOR}&\_&\mathsf{tt}&=&\mathsf{tt} \\
      \mathsf{parallelOR}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathsf{parallelOR}&\bot&\bot&=&\bot
    \end{array}
  \end{displaymath}
  For the input-output pair $(\mathsf{tt},\mathsf{tt}),\mathsf{tt}$,
  there is no one minimal input that produces this output. We have
  both $\mathrm{parallelOR}(\mathsf{tt},\bot) = \mathsf{tt}$ and
  $\mathrm{parallelOR}(\bot,\mathsf{tt}) = \mathsf{tt}$, which are
  incomparable and their greatest lower bound $(\bot,\bot)$ gives the
  output $\bot$.

  Parallel OR is famous because it is not \emph{sequential}, meaning
  intuitively that it cannot be implemented without running the two
  arguments in parallel to see if one of them returns
  $\mathsf{tt}$. That fact that it exists in the standard domain
  theoretic semantics of PCF means that this semantics is not fully
  abstract. Since Parallel OR is not stable, one might hope that
  stability is enough to capture sequentiality, and hence potentially
  give a fully abstract model of PCF. However, the following ternary
  function $\mathbb{B}_\bot^3 \to \{\top,\bot\}$ is stable but admits
  no sequential implementation that fixes an order that the arguments
  are examined in:
  \begin{displaymath}
    \begin{array}{l@{(}l@{,~}l@{,~}l@{)~}c@{~}l}
      \mathrm{gustave}&\mathsf{tt}&\mathsf{ff}&\_&=&\top \\
      \mathrm{gustave}&\mathsf{ff}&\_&\mathsf{tt}&=&\top \\
      \mathrm{gustave}&\_&\mathsf{tt}&\mathsf{ff}&=&\top \\
      \mathrm{gustave}&\_&\_&\_&=&\bot
    \end{array}
  \end{displaymath}
  Despite there being no one minimal input that achieves the output
  $\top$, each of the minimal inputs that can achieve this output are
  pairwise incomparable, so for each specific input that gets output
  $\top$ there is a unique minimal input that achieves it (listed in
  the first three lines of the definition).

  In terms of program slicing, however, the $\mathrm{gustave}$
  function does not present a problem. For any particular run of the
  program, there is an unambiguous minimal input that achieves the
  output, no matter that it was not achieved by a sequential
  processing of the input.
\end{example}

The definition of stability has a nice operational meaning, but seems
a little ad-hoc. We can rephrase stability in terms of Galois
connections / adjunctions between principal downsets / slices of the
partial orders.

@@

\begin{proof}
  If $f$ is stable, then define a left adjoint
  $g_x : \downset{f(x)} \to \downset{x}$ by setting $g_x(y)$ to be the
  minimal $x_0$ required by stability. This is monotone: if
  $y \leq y'$, then we know that $y \leq y' \leq f(g_x(y'))$ by
  definition of $g_x$, so $g_x(y) \leq g_x(y')$ by minimality of
  $g_x(y)$. For the adjointness, let $x' \leq x$ and $y \leq
  f(x)$. Then if $g_x(y) \leq x'$, we have
  $y \leq f(g_x(y)) \leq f(x')$ by monotonicity of $f$ and the first
  part of stability. In the other direction, if we have
  $y \leq f(x')$, then by uniqueness we have $g_x(y) \leq x'$.

  If, for every $x$, $f_x$ has a left adjoint $g_x$, then for any
  $x', y$ we have $y \leq f_x(x') \Leftrightarrow g_x(y) \leq x'$. So
  $g_x(y)$ is the element that satisfies $y \leq f(g_x(y))$, and it is
  minimal since if $y \leq f_x(x'_0)$ then $g_x(y) \leq x'_0$.
\end{proof}
